% Chapter 5: RL in Games

Computing Nash equilibria in extensive-form games with imperfect information is PPAD-complete \citep{Daskalakis2009}.\footnote{PPAD (Polynomial Parity Arguments on Directed graphs) is a complexity class for problems where existence is guaranteed by a parity argument but no efficient algorithm is known. PPAD-completeness implies no polynomial-time algorithm exists unless PPAD = P.} Counterfactual regret minimization (CFR) circumvents worst-case hardness by exploiting game structure, achieving $O(1/\sqrt{T})$ convergence to $\varepsilon$-Nash equilibrium in two-player zero-sum games \citep{zinkevich2008}. This section applies CFR to the durable goods monopoly problem, benchmarking against the theoretical results of \citet{gul1986foundations} and \citet{ausubel1989reputation}.

\subsection{Counterfactual Regret Minimization}

An extensive-form game consists of a game tree with information sets $\mathcal{I}_i$ partitioning player $i$'s decision nodes. A behavioral strategy $\sigma_i: \mathcal{I}_i \to \Delta(\mathcal{A})$ assigns action probabilities at each information set. The counterfactual value of action $a$ at information set $I$ is
\begin{equation}
v^\sigma_i(I, a) = \sum_{h \in I} \pi^\sigma_{-i}(h) \sum_{z \sqsupseteq ha} \pi^\sigma(ha, z) u_i(z)
\end{equation}
where $\pi^\sigma_{-i}(h)$ is the probability opponents reach $h$, and $\pi^\sigma(ha, z)$ is the probability of reaching terminal $z$ from $ha$.\footnote{The counterfactual formulation weights by opponent reach $\pi^\sigma_{-i}(h)$ rather than joint reach $\pi^\sigma(h)$, ensuring non-zero updates even at rarely-visited information sets. This avoids the vanishing gradient problem that affects direct policy gradient methods in extensive-form games.} Cumulative regret for action $a$ is $R^T(I,a) = \sum_{t=1}^T [v^{\sigma^t}(I,a) - v^{\sigma^t}(I)]$. CFR updates via regret matching:
\begin{equation}
\sigma^{T+1}(I, a) = \frac{[R^T(I, a)]^+}{\sum_{a'} [R^T(I, a')]^+}, \quad [x]^+ = \max(x,0).
\end{equation}
The average strategy $\bar{\sigma}^T$ converges to $\varepsilon$-Nash with $\varepsilon = O(|\mathcal{I}|\sqrt{|\mathcal{A}|}\Delta / \sqrt{T})$ \citep{zinkevich2008}.\footnote{CFR+ \citep{tammelin2014solving} modifies the update to $R^{T+1}(I, a) = [R^T(I, a) + r^{T+1}(I, a)]^+$ and weights iteration $t$ by $t$. These modifications improve practical convergence without changing the asymptotic rate.}

\subsection{Neural Extensions}

Tabular CFR stores regrets at every information set, infeasible when $|\mathcal{I}| > 10^{14}$. Two neural approaches scale to large games.

\subsubsection{Deep CFR}

\citet{Brown2019} approximate cumulative regrets with a neural network $V_\theta: \mathcal{I} \times \mathcal{A} \to \mathbb{R}$ trained on sampled (information set, action, regret) tuples:
\begin{equation}
L_V(\theta) = \mathbb{E}_{(I, a, r) \sim \mathcal{M}} \left[ (V_\theta(I, a) - r)^2 \right].
\end{equation}
A separate network $\Pi_\phi$ approximates the average strategy via KL-divergence minimization. The current strategy derives from regret matching: $\sigma(I, a) \propto [V_\theta(I, a)]^+$.

\subsubsection{Neural Fictitious Self-Play}

NFSP \citep{heinrich2016deep} combines fictitious play \citep{Brown1951} with deep Q-learning. Each player maintains two networks: a best-response network $Q_\theta$ trained via DQN, and an average strategy network $\bar{\pi}_\phi$ trained via supervised learning on the best-response actions:
\begin{equation}
L_{\text{RL}}(\theta) = \mathbb{E}\left[ \left( r + \gamma \max_{a'} Q_{\theta^-}(I', a') - Q_\theta(I, a) \right)^2 \right], \quad
L_{\text{SL}}(\phi) = \mathbb{E}\left[ -\log \bar{\pi}_\phi(a | I) \right].
\end{equation}
During play, agents follow $\bar{\pi}_\phi$ with probability $1-\eta$ and $Q_\theta$ with probability $\eta$.

\subsubsection{Poker Results}

These methods achieved superhuman performance in poker. Deep CFR attained exploitability of 0.036 big blinds per hand in heads-up limit hold'em \citep{Brown2019}. Libratus defeated top human professionals in heads-up no-limit hold'em using nested subgame solving with CFR \citep{brown2017superhuman}. Pluribus extended to six-player no-limit hold'em \citep{brown2019superhuman}. The game tree of heads-up no-limit hold'em contains $\sim 10^{161}$ states; these results demonstrate that CFR-based methods scale to economically relevant game sizes.

\subsection{The Coase Conjecture}

\citet{coase1972durability} conjectured that a durable goods monopolist loses market power when buyers are patient: unable to commit to future prices, the seller competes with her future self, eroding rents. \citet{stokey1981rational} showed that a monopolist facing rational consumers cannot price discriminate intertemporally; \citet{bulow1982durable} proved that in the ``no-gap'' case (cost above lowest valuation), the monopolist prefers renting to selling. \citet{gul1986foundations} provided the definitive formalization: with a continuum of buyer types and continuous time, as the inter-offer interval $\Delta t \to 0$, price converges to marginal cost instantaneously and all trade occurs at $t = 0$. The ``gap case'' (cost below lowest valuation) admits a unique stationary equilibrium \citep{ausubel1989reputation, ausubel2002bargaining}. We benchmark against this case.

\subsubsection{Model}

A seller with zero cost faces a buyer with private valuation $v \in \{v_L, v_H\}$, where $\Pr(v = v_H) = \pi$. In each period $t = 1, 2, \ldots$, the seller posts price $p_t$; the buyer accepts or rejects. Payoffs upon acceptance at $t$: seller receives $\delta^{t-1} p_t$, buyer receives $\delta^{t-1}(v - p_t)$, where $\delta \in (0,1)$ is the common discount factor. The game ends upon acceptance or after $T$ periods.

This is the ``gap case'' of \citet{ausubel2002bargaining}: seller cost (0) is strictly below the lowest buyer valuation ($v_L$), guaranteeing trade in equilibrium. The gap case admits a unique stationary equilibrium.

\subsubsection{Theoretical Benchmark}

The screening price $P^*(\delta)$ makes the high-type buyer indifferent between accepting now and waiting:
\begin{equation}
v_H - P^* = \delta(v_H - v_L) \implies P^*(\delta) = v_H - \delta(v_H - v_L).
\end{equation}
The seller's optimal strategy depends on $\pi$. Let $\Pi_{\text{screen}} = \pi P^* + (1-\pi)\delta v_L$ and $\Pi_{\text{pool}} = v_L$. The seller screens if $\Pi_{\text{screen}} > \Pi_{\text{pool}}$, yielding threshold
\begin{equation}
\pi^* = \frac{v_L(1-\delta)}{P^* - \delta v_L} = \frac{1}{2} \quad \text{when } v_H = 2v_L.
\end{equation}
For $\pi < \pi^*$: pooling equilibrium (offer $v_L$ immediately). For $\pi > \pi^*$: screening equilibrium (offer $P^*$, then $v_L$ if rejected).

The Coase conjecture manifests as $\delta \to 1$: the screening price $P^*(\delta) \to v_L$, so the seller cannot extract surplus from high types.

\subsubsection{Computational Results}

We model the bargaining game as an extensive-form game and apply CFR. Parameters: $v_L = 100$, $v_H = 200$, $T = 2$ periods. The seller's information set is the rejection history; the buyer's information set includes her private type.

\begin{table}[htbp]
\centering
\caption{CFR Equilibrium vs. Theory: $\pi$-sweep at $\delta = 0.5$}
\label{tab:coase}
\input{../ch05_rl_in_games/sims/durable_goods_results}
\begin{minipage}{0.9\textwidth}
\footnotesize
Notes: P(Screen) is the probability the seller offers $P^* = 150$ in period 1. Theory column gives the predicted probability (0 for pooling, 1 for screening). NashConv is the sum of exploitabilities. 5,000 CFR iterations per configuration.
\end{minipage}
\end{table}

Table \ref{tab:coase} reports results from varying $\pi$ at fixed $\delta = 0.5$. The algorithm recovers the phase transition at $\pi^* = 0.5$: for $\pi < 0.5$, the seller pools with probability approaching 1; for $\pi > 0.5$, the seller screens with probability approaching 1. The transition is sharp.

A $\delta$-sweep at $\pi = 0.7$ confirms the screening price formula: $P^*(\delta) = 200 - 100\delta$ with zero error across $\delta \in [0.1, 0.9]$. As $\delta$ increases, screening becomes unprofitable (Coase conjecture): at $\delta = 0.75$, the seller switches to pooling.

\subsubsection{Validation}

Four stress tests verify the results.\footnote{(1) Awkward primes $(v_L, v_H) = (37, 83)$: converged to $P^* = 55$ (theory: 55.4). (2) Information leak: single seller information set at root. (3) Grid shift with 150 removed: used 145 (99.4\%). (4) 3-period game: found $P_1 = 120$ vs.\ theory 136.} The 3-period result merits explanation. At $P_1 = 136$, the high buyer is exactly indifferent: surplus $= v_H - 136 = 64 = \delta(v_H - 120)$. If the buyer randomizes, the seller prefers $P_1 = 120$ (revenue 108 vs.\ 86.4 if buyer rejects 136). CFR found the equilibrium robust to tie-breaking.

\subsection{Discussion}

CFR computed equilibria matching the theoretical predictions of \citet{gul1986foundations} without encoding economic structure. The algorithm discovered that: (1) the seller screens when $\pi$ is high, pools when $\pi$ is low, with threshold at $\pi^* = 0.5$; (2) the screening price satisfies the high-buyer indifference condition; (3) as $\delta \to 1$, screening collapses (Coase conjecture). These properties emerged from regret minimization alone.

The poker-bargaining correspondence is exact: buyer valuation maps to hole cards, price acceptance to calling a bet, rejection to folding. Both involve signaling private information through sequential actions. The success of poker algorithms on this benchmark suggests applicability to other bargaining and mechanism design problems with analytically intractable equilibria.
