Economic decisions rarely occur in isolation. Rather, they unfold sequentially over time, with choices today influencing opportunities tomorrow. They also interact with the decisions of others. The temporal and strategic dimension of economic behavior makes it difficult to model mathematically. How can we understand and predict decisions in changing environments, evolving strategic competitors, and varying policy rules? Two methodological fields have grown to tackle this challenge: structural econometrics in economics and reinforcement learning in computer science. Although these fields have developed independently, they address similar problems through complementary approaches.

Structural econometrics, rooted in economic theory, interprets observed choices as optimal decisions made under institutional and technological constraints. By estimating the underlying preference and technology parameters driving these choices, econometricians can perform counterfactual analysis—predicting how agents would behave if the "rules of the game" changed. This ability to simulate policy counterfactuals makes structural models invaluable for economic policy design and evaluation. However, structural approaches have traditionally been limited by computational constraints in handling high-dimensional state spaces and strategic interactions.

Reinforcement learning (RL), meanwhile, has emerged as a powerful paradigm for sequential decision-making under uncertainty. RL generalizes dynamic programming by learning optimal policies directly from interaction with an environment, without requiring explicit specification of transition probabilities or reward functions. Recent advances in deep reinforcement learning have demonstrated remarkable successes from game play to robotics, language model alignment, and logistics optimization. These methods excel at finding approximately optimal policies in high-dimensional environments where traditional dynamic programming would be intractable.

Both fields evolved through parallel yet independent trajectories. Structural econometrics emerged from the Cowles Commission's work in the 1940s, establishing frameworks for connecting economic theory with statistical inference. It progressed through McFadden's discrete choice models in the 1970s and Rust's dynamic programming approaches in the 1980s to modern computational methods. Reinforcement learning traces its origins to early AI pioneers like Shannon and Minsky, crystallizing as a field through Sutton and Barto's unifying work on temporal difference learning in the 1980s. Both disciplines experienced transformative computational advances in the 2010s—structural econometrics incorporating machine learning techniques for high-dimensional problems and reinforcement learning achieving breakthroughs with deep neural networks.

Methodologically, all dynamic programming and reinforcement learning algorithms can be subsumed in a single framework. Reinforcement learning algorithms can be understood as a generalization of dynamic programming that relaxes the requirement of knowing the transition probabilities and reward functions. This makes convergence slower, as model structure cannot be exploited, but allows RL to scale-up and overcome the curse of dimensionality that plagues DP. This helps us to understand why RL methods, under sufficient exploration of the state space, converge to the same optimal policies identified by traditional dynamic programming. We offer a few concrete examples to show how tabular and deep reinforcement learning can mitigate the curse of dimensionality in dynamic optimization problems and dynamic games. 

We then examine how reinforcement learning has been applied to solve structural economic models. We look at experience-based equilibria in Asker et al. (2020), demonstrate how reinforcement learning principles can be used to define and compute equilibrium concepts in games with asymmetric information. In industrial organization, RL techniques have proven valuable for analyzing dynamic oligopoly models with endogenous innovation, entry/exit, and mergers—problems that were previously computationally prohibitive. We also explore applications in dynamic discrete choice modeling, where policy gradient methods combined with indirect inference provide new estimation approaches for models with continuous state variables and unobserved heterogeneity. In macroeconomics, deep reinforcement learning has enabled solution of general equilibrium models without relying on simplifying approximations.

The paper further investigates Inverse Reinforcement Learning (IRL)—a subfield that seeks to recover reward functions from observed behavior. We highlight the strong methodological parallels between IRL and structural estimation in economics. Both aim to infer underlying preferences from choice data, but approach the problem through different computational techniques. Maximum entropy IRL, in particular, shares deep connections with discrete choice models in econometrics, with the Boltzmann policy formulation corresponding precisely to multinomial logit specifications with Gumbel-distributed utility shocks. Recent innovations like Sharma et al.'s (2018) adaptation of Hotz-Miller's conditional choice probability approach demonstrate how econometric methods can enhance IRL's computational efficiency, while IRL's flexible function approximation capabilities can benefit structural estimation of complex preferences.

Reinforcement learning in strategic settings forms another critical area of overlap. We examine how multi-agent reinforcement learning connects to evolutionary game theory through replicator dynamics, and review algorithms specifically designed for competitive environments. Methods like counterfactual regret minimization that have achieved superhuman performance in games offer powerful new approaches for modeling strategic interaction in economics, with applications from auction design to industrial organization and macroeconomic policy.

Economic theory provides valuable structure that can significantly enhance reinforcement learning algorithms. By incorporating domain-specific knowledge like monotonicity in demand curves, unimodality in auction revenues, and low-rank structures in multi-product markets, reinforcement learning can achieve exponentially improved learning efficiency. For dynamic pricing problems, economic structure enables algorithms to achieve logarithmic rather than square-root regret. In resource-constrained optimization settings, economic principles of marginal value versus marginal cost naturally map to dual formulations that balance exploration and exploitation. When modeling strategic interactions, economic equilibrium concepts provide stability guarantees and solution refinements. These structural assumptions not only accelerate learning but also produce more interpretable and generalizable policies, creating a bridge between theory-driven and data-driven approaches to decision-making.

Finally, we look at real world applications of SE and RL where these methodological innovations have delivered practical impact. We examine how structural econometric models have shaped antitrust policy, market design, program evaluation, and monetary policy, while reinforcement learning has begun to be applied to financial trading, energy management, transportation systems, recommendation systems, and marketing optimization. 

Throughout this exploration, several common themes emerge. Both fields grapple with similar fundamental challenges: the curse of dimensionality in large state spaces, equilibrium multiplicity in strategic settings, and identification of underlying preferences from choice data. For economists, reinforcement learning offers powerful computational tools to solve economic models. For computer scientists, economic theory provides principled structural assumptions that can accelerate learning, reduce sample complexity, enable identification, and enhance interpretability. 

Our survey and interpretative essay seek to contribute in a couple of ways. Firstly, it seeks to provide a comprehensive update of relevant papers at the intersection of SE and RL. Most surveys between economics and computer science have focused on supervised learning (Athey and Imbens 2019). This survey differs from Charpentier et al 2021 by focusing on the methodological connections. It builds on Iskhakov et al. 2021, but covers many new topics like inverse reinforcement learning, using RL to solve equilibria of games, and using of SE to improve RL. 


The remainder of this paper proceeds as follows. Section 2 traces the historical development of both fields. Section 3 establishes the connection between dynamic programming and reinforcement learning and offers concrete examples. Section 4 examines applications of reinforcement learning to solving structural economic models. Section 5 explores inverse reinforcement learning and its connections to structural estimation. Section 6 investigates reinforcement learning in game-theoretic settings, replicator dynamics and scalability. Section 7 discusses how structural assumptions can enhance online reinforcement learning or bandits. Section 8 surveys real-world applications across various domains. We conclude by identifying promising directions for future research at the interdisciplinary frontier.